Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
finished an editing pass over section 3.2

MRFy's search computes a sequence of \emph{generations}.
A~generation is a collection of scored {placements}.
We~pass the
search function a \texttt{Scorer}, which runs Viterbi's algorithm.
\smallverbatiminput{scoredecl}
A~search heuristic is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallfuzzverbatiminput{2.3pt}{strategy}
In this record,
\begin{itemize}
\item
Function \texttt{gen0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may be no better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} function considers a list of type
\texttt{[Scored~Age]}.
At~the head of this list is the score of the best
placement of the trial generation, together with the \texttt{Age} of the trial
generation, which is the total number of generations tried.
The remaining elements of the list hold the history of scores and ages for
previously \emph{accepted} generations.
(The list omits generations that didn't survive the \texttt{accept} trial.)
For convenience, the \texttt{accept} function also receives
the \texttt{Age} of the trial generation.
\item
Finally, the \texttt{quit} function decides when to stop the search.
It~too considers the list of accepted generations and the age of the current
generation. 
\end{itemize}
This interface provides ample leeway for experiments.
For~example, random-mutation hill climbing accepts only generations
with better scores than their predecessors,
while simulated annealing may accept a generation with a worse score.
A~genetic algorithm uses \texttt{nextGen} to
combine placements from the previous generation, while
random-mutation
hill climbing simply permutes a placement into a new one.
The \texttt{quit} interface makes it easy
to implement such termination criteria as a fixed number of generations
or a limit of going $k$~generations without achieving more than $E$
improvement in score.

\seclabel{search}

We hope you find the search simple and straightforward:
\smallfuzzverbatiminput{11pt}{search}

We found it easy to write several stochastic search heuristics thanks to
this framework.
We first implemented random-mutation hill climbing in one day and roughly
fifty lines of code.
We next implemented simulated annealing, and thanks to Haskell's module system,
we imported the \texttt{nextGen}, \texttt{gen0}, and \texttt{quit} functions
from the random-mutation hill climbing module.
Indeed, we had only to write a new \texttt{accept} function, comprising only ten
lines of code and half an hour of our time.
We then implemented a genetic algorithm approach; this was still able to import
existing \texttt{gen0} and \texttt{quit} functions, and its \texttt{accept}
function was only a few lines of code.
Understandably, though, the complexities of representing ``crossover'' of parent
placements required forty lines of code and much of a day of thought.

>>>>

MRFy's search computes a sequence of \emph{generations}.
A~generation is a collection of scored {placements}.
We~pass the
search function a \texttt{Scorer}, which runs Viterbi's algorithm.
\smallverbatiminput{scoredecl}
A~search heuristic is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallfuzzverbatiminput{2.3pt}{strategy}
In this record,
\begin{itemize}
\item
Function \texttt{gen0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may be no better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} function considers a list of type
\texttt{[Scored~Age]}.
At~the head of this list is the score of the best
placement of the trial generation, together with the \texttt{Age} of the trial
generation, which is the total number of generations tried.
The remaining elements of the list hold the history of scores and ages for
previously \emph{accepted} generations.
(The list omits generations that didn't survive the \texttt{accept} trial.)
For convenience, the \texttt{accept} function also receives
the \texttt{Age} of the trial generation.
\item
Finally, the \texttt{quit} function decides when to stop the search.
It~too considers the list of accepted generations and the age of the current
generation. 
\end{itemize}
\seclabel{search}%
We hope you find the search simple and straightforward:
\smallfuzzverbatiminput{11pt}{search}

The \texttt{SearchStrategy} interface provides ample scope for
experiments. 
Random-mutation hill took fifty lines of code and one~day to
implement.
Simulated annealing reuses
 \texttt{nextGen}, \texttt{gen0}, and \texttt{quit} 
from random-mutation hill climbing.
Simulated annealing's  new \texttt{accept} function took ten
lines of code and half an hour to implement.
(Hill climbing accepts a generation only if it scores better than its
predecessor;
simulated annealing may accept a generation with a worse score.)
Our genetic algorithm reuses the same
\texttt{gen0} and \texttt{quit} functions, and its \texttt{accept}
function was only a few lines of code.
But~the recombination of parent
placements in a new \texttt{nextGen} function took forty lines of code
and a full day to implement.
Our~current \texttt{quit} function simply runs for a fixed number of
generations, but its interface will enable us to try other criteria,
such as stopping once $k$~generations have passed without more than $E$
improvement in score.
